### 2024-09-25 teaching/math240f24/draft.tex
transfer to dorothea2

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}

@@

@@

A set of numbers near and dear to many mathematicians' hearts is the set of {\it prime} numbers $P$,
defined by
$$P = \bigl\{ p\in \NN : p\ge 2,\ \hbox{and if $p = ab$ then $\{a,b\} = \{1,p\}$}\bigr\}.$$
(You may have met a different definition of prime numbers in the past. Pause a moment and convince
yourself that the statement above defines the set of prime numbers as you know it.)

@@

\medskip\boldlabel\datestamp{10}{ix} Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.
But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
suppose your friend says: ``If it snows tomorrow I'll work in Trottier with you.'' If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

@@

Now we practise translating converting mathematical statements written in English into formulas
in predicate logic. Suppose we want to write: ``Every integer is even or odd.'' The universe here is
is the set $\ZZ$ of integers. The word ``every'' has the same meaning as ``for all,'' so right
off the bat, we can reexpress the statement as: ``For all $n\in \ZZ$, $n$ is even or $n$ is odd.''
In symbols, this is
$$ \forall n\,(n\ \hbox{even} \vee n\ \hbox{odd}).$$
Lastly, we need to figure out how to express the property of being even or being odd.
An integer $n$ is even if an only if it is a multiple of two; that is, if there is some integer $k$
such that $2k = n$. Likewise, an integer is odd if and only if it is one more than a multiple of two.
The corresponding formula is $\exists k : 2k+1 = n$.
So our statement can be expressed
$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists k : n = 2k+1)\bigr).$$
The variable $k$ appears twice in this formula, but its first instance is independent of its second instance,
because of the parentheses. (Readers who write computer programs will be familiar with the concept of
a variable ``going out of scope.'') So there is nothing wrong with this formula, but to be extra
clear that the first $k$ is different from the second $k$, why don't we replace it with a different letter?
Thus we arrive at
$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists l : n = 2l+1)\bigr),$$
a formula in predicate logic that means ``every integer is even or odd.''

@@

In the following proof, we use the fact that every nonempty subset of the natural numbers has
a least element, the proof of which belongs to a higher-level set theory course.

@@

Here are some examples and non-examples.
\medskip
\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
when written in set-builder notation.
\smallskip
\item{ii)} On the other hand, the set
$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
there is no element of $g$ with $-1$ as its first coordinate.
\smallskip
\item{iii)} If $X = \{1,2,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
is a function from $X$ to $Y$.
\smallskip
\item{iv)} The set
$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
is not a function, since there is no $y$ such that $(0,y)\in h$. However
if we amend the domain and consider
$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
\medskip
\bye

@@

\proof We shall construct a function $f:E\to F$. Fix one specific $x\in X$;
we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
$$f(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
Note that since $|A|$ is even for all $A\in E$, the cardinality of $f(A)$ is odd (in the first
case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
do by showing that $f$ is an inverse (as an exercise, you might try to prove bijectivity from
the definitions of injective and surjective).

@@

\proof We present the proof as a (difficult) exercise. Here is the roadmap.
Call $b\in B$ {\it unattached} if there is no $a\in A$ such that $f(a) = b$.
Let $h:B\to B$ be given by $h(b)  = f\bigl(g(b)\bigr)$. Given $b,b'\in B$, we shall say that
$b$ is a {\it peer} of $b'$ if either $b = b'$ or there exists some $n\in \NN$ such
that
$$ b = \underbrace{h(h(\cdots(h}_{n\ \hbox{\sevenrm times.}}(b'))\cdots))$$
Say that $b\in B$ is a {\it PAE} if it is the peer of an unattached element. (So unattached
elements are automatically PAEs, by setting $n=0$.)
\medskip
\item{a)} Show that if $a\in A$ is such that
$f(a)$ is a PAE, then
there is a unique element $b^*\in B$ such that $g(b^*) = a$, and that
this element is a PAE.
\medskip
\noindent By part (a), if $f(a)$ is a PAE, it makes sense to speak of $g^{-1}(a)$. It is
the element $b^*\in B$ such that $g(b^*) = a$. From here we define
$$ r(a) = \cases{ g^{-1}(a), & if $f(a)$ is a PAE;\cr
f(a), & otherwise.}$$
\medskip
\item{b)} Show that if $b\in B$ is a PAE
then so is $f\bigl(g(b)\bigr)$.
\item{c)} Show that $r$ is surjective. [{\it Hint}: Do a proof by cases. Every $b\in B$ is either
a PAE or it is not a PAE.]
\smallskip
\item{d)} We already proved in part (a) that if $f(a)$ is a PAE, then so is $r(a)$. Prove
the converse of this statement.
\smallskip
\item{e)} Show that $r$ is injective, and therefore bijective. [{\it Hint}:
Assume $r(a_1) = r(a_2)$, and do a proof by cases again. Part (d) will be useful here.]\slug
\medskip

Once again, this proof is outside the scope of the course. Don't lose sleep over it if you can't do it.
Feel free to email me with questions if you get stuck.

@@

\proof
Enumerate $A = \{a_0, a_2, \ldots\}$ and $B = \{b_0, b_2, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = 2^i3^j.$$
We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, f(b_{j'}).$
Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
Theorem of Arithmetic, we see that $i = i'$ and $j=j'$.

@@

\proclaim Theorem. The set $\QQ$ of rational numbers is countable.

@@

\datestamp{10}{} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
\medskip
\item{i)} The set $< = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
write $(2,\pi)\in <$, or $2<\pi$ (which is the more common notation for this relation).

>>>>

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
\def\divides{\mid}

\vglue24pt

@@

\vskip60pt

\vskip \medskipamount \hbox to\hsize {\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
%\smallskip
%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
%\smallskip
%\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}

\vfill\eject
\begingroup\headline{\hfil}\footline{\hfil}
\vglue160pt
\centerline{\titlefont I. FOUNDATIONS}
\vskip220pt
\bigskip
\begingroup\obeylines\eightssi
\hfill Die Mathematik ist in ihrer Entwickelung v\"ollig frei
\hfill und nur an die selbstredende R\"ucksicht gebunden,
\hfill dass ihre Begriffe sowohl in sich widerspruchslos sind,
\hfill als auch in festen durch Definitionen geordeneten Beziehungen
\hfill zu den vorher gebildeten,
\hfill bereits vorhandenen und bew\"ahrten Begriffen stehen.
\eightss
\smallskip
\hfill --- GEORG CANTOR, {\eightssi Grundlagen einer allgemeinen Mannigfaltigkeits­lehre} (1883)
\endgroup%\obeylines
\bigskip\goodbreak
\vfill\eject

@@

A set of numbers near and dear to many mathematicians' hearts is the set of {\it prime} numbers $P$,
defined by
$$P = \bigl\{ p\in \NN : p\ge 2,\ \hbox{and if $p = ab$ then $\{a,b\} = \{1,p\}$}\bigr\}.$$
(You may have met a different definition of prime numbers in the past. Pause a moment and convince
yourself that the statement above defines the set of prime numbers as you know it.)
\endgroup%\headline{\hfil}\footline{\hfil}

@@

\medskip\boldlabel\datestamp{10}{ix} Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.
But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
suppose your friend says, If it snows tomorrow I'll work in Trottier with you. If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

@@

Now we practise translating converting mathematical statements written in English into formulas
in predicate logic. Suppose we want to write, Every integer is even or odd. The universe here is
is the set $\ZZ$ of integers. The word ``every'' has the same meaning as ``for all,'' so right
off the bat, we can reexpress the statement as, For all $n\in \ZZ$, $n$ is even or $n$ is odd.
In symbols, this is
$$ \forall n\,(n\ \hbox{even} \vee n\ \hbox{odd}).$$
Lastly, we need to figure out how to express the property of being even or being odd.
An integer $n$ is even if an only if it is a multiple of two; that is, if there is some integer $k$
such that $2k = n$. Likewise, an integer is odd if and only if it is one more than a multiple of two.
The corresponding formula is $\exists k : 2k+1 = n$.
So our statement can be expressed
$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists k : n = 2k+1)\bigr).$$
The variable $k$ appears twice in this formula, but its first instance is independent of its second instance,
because of the parentheses. (Readers who write computer programs will be familiar with the concept of
a variable ``going out of scope.'') So there is nothing wrong with this formula, but to be extra
clear that the first $k$ is different from the second $k$, why don't we replace it with a different letter?
Thus we arrive at
$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists l : n = 2l+1)\bigr),$$
a formula in predicate logic that means, Every integer is even or odd.

@@

In the following proof, we use the {\it well-ordering principle}, which is the
fact that every nonempty subset of the natural numbers has
a least element. You can see the proof of the well-ordering principle in a higher-level set theory course.

@@

Here are some examples and non-examples.
\medskip
\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
when written in set-builder notation.
\smallskip
\item{ii)} On the other hand, the set
$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
there is no element of $g$ with $-1$ as its first coordinate.
\smallskip
\item{iii)} If $X = \{1,2,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
is a function from $X$ to $Y$.
\smallskip
\item{iv)} The set
$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
is not a function, since there is no $y$ such that $(0,y)\in h$. However
if we amend the domain and consider
$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
\medskip

@@

\proof We shall construct a function $f:E\to F$. Fix one specific $x\in X$;
we can do this because $X\ne \emptyset$. Now, for all $A\in E$, let
$$f(A) = \cases{A\setminus\{x\}, & if $x\in A$;\cr A\cup \{x\}, & if $x\notin A$.\cr}$$
Note that since $|A|$ is even for all $A\in E$, the cardinality of $f(A)$ is odd (in the first
case it is $|A|-1$ and in the second case it is $|A|+1$. This shows that $f$ is indeed a function
with codomain $F$. Now to prove $|E| = |F|$ we will show that $f$ is bijective, which we
do by showing that $f$ is an inverse (as an exercise, you might instead try to prove bijectivity from
the definitions of injective and surjective).

@@

\noindent\llap{*}{\it Proof.}\enspace We present the proof as a (difficult) exercise. Here is the roadmap.
Call $b\in B$ {\it unattached} if there is no $a\in A$ such that $f(a) = b$.
Let $h:B\to B$ be given by $h(b)  = f\bigl(g(b)\bigr)$. Given $b,b'\in B$, we shall say that
$b$ is a {\it peer} of $b'$ if either $b = b'$ or there exists some $n\in \NN$ such
that
$$ b = \underbrace{h(h(\cdots(h}_{n\ \hbox{\sevenrm times.}}(b'))\cdots))$$
Say that $b\in B$ is a {\it PAE} if it is the peer of an unattached element. (So unattached
elements are automatically PAEs, by setting $n=0$.)
\medskip
\item{a)} Show that if $a\in A$ is such that
$f(a)$ is a PAE, then
there is a unique element $b^*\in B$ such that $g(b^*) = a$, and that
this element is a PAE.
\medskip
\noindent By part (a), if $f(a)$ is a PAE, it makes sense to speak of $g^{-1}(a)$. It is
the element $b^*\in B$ such that $g(b^*) = a$. From here we define
$$ r(a) = \cases{ g^{-1}(a), & if $f(a)$ is a PAE;\cr
f(a), & otherwise.}$$
\medskip
\item{b)} Show that if $b\in B$ is a PAE
then so is $f\bigl(g(b)\bigr)$.
\item{c)} Show that $r$ is surjective. [{\it Hint}: Do a proof by cases. Every $b\in B$ is either
a PAE or it is not a PAE.]
\smallskip
\item{d)} We already proved in part (a) that if $f(a)$ is a PAE, then so is $r(a)$. Prove
the converse of this statement.
\smallskip
\item{e)} Show that $r$ is injective, and therefore bijective. [{\it Hint}:
Assume $r(a_1) = r(a_2)$, and do a proof by cases again. Part (d) will be useful here.]\slug
\medskip

Once again, this proof is outside the scope of the course. Don't lose sleep over it if you can't do it.
Feel free to ask questions at office hours if you get stuck.

@@

\proof
Enumerate $A = \{a_0, a_1, \ldots\}$ and $B = \{b_0, b_1, \ldots\}$. Now given
$(a_i, b_j)\in A\times B$ for some $i,j\in \NN$, we can let
$$f(a_i, b_j) = 2^i3^j.$$
We show that this defines an injective function. Suppose that $f(a_i, b_j) = f(a_{i'}, f(b_{j'}).$
Then $2^i3^j = 2^{i'} 3^{j'}$, and by the uniqueness of prime factorisations in the Fundamental
Theorem of Arithmetic, we see that $i = i'$ and $j=j'$.

@@

\proclaim Theorem \advthm. The set $\QQ$ of rational numbers is countable.

@@

\datestamp{10}{} A {\it relation} on a set $X$ is a subset $R\subseteq X\times X$. If $(a,b)\in R$
we write $aRb$ and say ``$a$ is related to $b$.'' Here are some examples.
\medskip
\item{i)} The set $L = \bigl\{(a,b)\in \RR\times\RR : a<b\bigr\}$ is a relation. For instance, we can
write $(2,\pi)\in L$, or $2L\pi$, or $2<\pi$ (which is the more common notation for this relation).
\smallskip
\item{ii)} The set $E = \bigl\{ (a,b) \in \RR\times \RR : a = b\bigr\}$ is a relation (it is the ``equals''
relation).
\smallskip
\item{iii)} On $\ZZ$, the set
$$R = \bigl\{(-1, 4), (8, -3), (0, 0), (0,1)\bigr\}$$
is a relation.
\smallskip
\item{iv)} For any set $A$, a function $f:A\to A$ is by definition a subset of $A\times A$, and hence
is an example of a relation.
\smallskip
\item{v)} Let $H$ be the set of all humans, define $M\subseteq H\times H$ by setting
$(h_1, h_2)$ if and only if $h_1$ is married to $h_2$. For instance,
$$(\hbox{Michelle Obama}, \hbox{Barack Obama})\in M.$$
On the other hand, it is unfortunately the case that
$$(\hbox{Marcel Goh}, \hbox{Taylor Swift})\notin M.$$
\medskip

\medskip\boldlabel Properties of relations. Let $R\subseteq A\times A$ be a relation. $R$ is called
\medskip
\item{i)} {\it reflexive} if for all $a\in A$, $aRa$;
\smallskip
\item{ii)} {\it symmetric} if for all $a,b\in A$ with $aRb$, we also have $bRa$; and
\smallskip
\item{iii)} {\it transitive} if for all $a,b,c\in A$ with $aRb$ and $bRc$, we also have $aRc$.
\medskip
Going back to the examples we had above, the relation $L$ is transitive but neither reflexive nor symmetric,
the relation $E$ satisfies all three properties, the relation $R$ satisfies none of them, and the relation
$M$ is symmetric but neither reflexive nor transitive. (Convince yourself of all of these facts.)

As a more involved example, let $X$ be any set with $|X|\ge 2$. On $2^X$, define a relation $R$ by
$$(A,B)\in R \Leftrightarrow A\cap B\ne \emptyset$$
for all $A,B\subseteq X$. Which of the three properties above does $R$ satisfy?

Is it reflexive? Well, is it true that for all $A\subseteq X$, $A\cap A \ne \emptyset$? The answer
is no, since we have $\emptyset \cap \emptyset = \emptyset$, so we have $(\emptyset, \emptyset)\notin R$.

Is $R$ symmetric? Well, if $A\cap B \ne \emptyset$, then since $\cap$ is commutative, we have
$B\cap A \ne \emptyset$ as well, so $(A,B)\in R$ implies that $(B,A)\in R$. In other words, yes, $R$ is symmetric.

Is $R$ transitive? If $A\cap B\ne \emptyset$ and $B\cap C\ne \emptyset$, does that necessarily mean
that $A\cap C\ne\emptyset$? The answer is no. Here's the proof. Since $|X|\ge 2$ we can find $x,y\in X$
with $x\ne y$. Let $A = \{x\}$, $B = \{x,y\}$, and $C = \{y\}$. Then $A\cap B = \{x\}\ne \emptyset$,
$B\cap C = \{y\}\ne \emptyset$, but alas $A\cap C = \emptyset$.

\medskip\boldlabel Equivalence relations and classes. If $R$ is reflexive, symmetric, and transitive, then we say
that $R$ is an {\it equivalence relation}. Here are two examples (as an exercise, prove that both of
these are equivalence relations).
\medskip
\item{i)} Let $F$ be the set of all formulas in propositional logic with the variables $p$, $q$, and $r$,
and operations $\neg$, $\wedge$, and $\vee$. The relation $\equiv$ defined by $f_1 \equiv f_2$
if and only if $f_1$ and $f_2$ have the same truth table is an equivalence relation.
\smallskip
\item{ii)} On the set $\RR^2$, define the relation $R \subseteq \RR^2\times \RR^2$ by letting
$(x,y) R (w,z)$ if and only if $\sqrt{x^2 + y^2} = \sqrt{w^2 + z^2}$. This is also an equivalence relation
\medskip
If $R$ is an equivalence relation, often we shall write $x\sim y$ to mean $xRy$. Sometimes we might
even just say that $\sim$ {\it is} the equivalence relation.

Let $R\subseteq A\times A$ be an equivalence relation. Define the {\it equivalence class} of $a\in A$
to be the set
$$[a] = \{b\in A : a\sim b\}.$$
This is the set of all $b\in A$ that are related to $a$.

Take for instance the example $F$ above of all formulas under the relation $\equiv$. The equivalence class
of the formula $p\vee\neg p$ is the set of all formulas whose truth table contains only $1$s, that is,
the set of all tautologies.
And for the example $R$ above, the
equivalence class of the point $(1,3)$ is the set
$$\bigl[ (1,3)\bigr] = \bigl\{ (x,y)\in \RR^2 : \sqrt\{10\} = \sqrt{x^2 + y^2}\bigr\};$$
that is, the circle of radius $\sqrt{10}$ centred about the origin.

We have the following proposition concerning equivalence relations.

\proclaim Proposition \advthm. Let $R$ be an equivalence relation on $A$. Then
\medskip
\item{i)} for all $x\in A$, $x\in [x]$;
\smallskip
\item{ii)} for all $x,y\in A$, $x\sim y$ if and only if $[x] = [y]$; and
\smallskip
\item{iii)} for all $x,y\in A$, $x\not\sim y$ if and only if $[x] \cap [y] =  \emptyset$.
\medskip

\proof Let $x\in A$. Since $R$ is reflexive, $x\sim x$, so we have $x\in [x]$. This proves (i).

For part (ii), first we prove the ``only if'' direction. Let $x,y\in A$ be such that $x\sim y$. To prove that
$[x]\subseteq [y]$, we let $z\in [x]$; so $x\sim z$. By symmetry, we have $z\sim x$, and this
combined with $x\sim y$ allow us to deduce that $z\sim y$, by transitivity. Then by symmetry again,
we have $y\sim z$, so $z\in [y]$. {\it Mutatis mutandis}, i.e., by swapping the roles of $x$ and $y$,
we also have $[y]\subseteq [x]$. Hence $[x] = [y]$.

Now for the ``if'' direction of (ii). Let $x,y\in A$ be such that
$[x] = [y]$. By (i), we have $x\in [x]$, but
then this means that $x\in [y]$. By definition of $[y]$, this means that $y\sim x$, so $x\sim y$
by symmetry.

On to part (iii). We prove both implications by contraposition (that is, we negate both sides of the
statement). Let $x,y\in A$ be such that $[x]\cap [y] \ne \emptyset$. This means there is some $z\in A$ such
that $z\in [x]$ and $z\in [y]$; so $x\sim z$ and $y\sim z$. By symmetry, $z\sim y$,
so by transitivity, we have $x\sim y$.

On the other hand, suppose that $x,y\in A$ satisfy $x\sim y$.
By (i) again, we have $[x] = [y]$, so $[x] \cap [y] = [x] \ne \emptyset$,
where we know that $[x]\ne \emptyset$ because it contains at least $x$.\slug

\noindent
\datestamp{11}{} \hskip\parindent Let $A$ be a set. A {\it partition} of $A$ is a set $P$ of subsets
of $A$ (i.e., $P\subseteq 2^A$), such that
\medskip
\item{i)} for all $x\in A$ there exists $S\in P$ such that $x\in S$;
\smallskip
\item{ii)} for all $S_1, S_2\in P$ with $S_1\ne S_2$, the intersection $S_1\cap S_2$ is empty; and
\smallskip
\item{iii)} $\emptyset\notin P$.
\medskip
Here are some examples.
\medskip
\item{i)} Let $E$ be the set of all even integers and $F$ the set of all odd integers. Then $\{E,F\}$
is a partition of $\ZZ$.
\smallskip
\item{ii)} The set $\bigl\{ (-\infty,0), \{0\}, (0,\infty)\bigr\}$ is a partition of $\RR$.
\medskip
If $\sim$ is an equivalence relation on a set $A$, then we can define the {\it quotient}
of $A$ by $\sim$ as the set of all equivalence classes of $A$ under $\sim$. We denote this set by
$$A/\sim = \bigl\{ [x] : x\in A\bigr\}.$$
We use the previous proposition to prove that equivalence relations give rise to partitions.

\proclaim Proposition \advthm. Let $A$ be a set and $\sim$ an equivalence relation on $A$.
Then $A/\sim$ is a partition of $A$.

\proof By part (i) the previous proposition, every $x\in A$ belongs to the equivalence class $[x]$.
Then, by part (ii) of the previous proposition, we know that for any equivalence classes
$[x]$ and $[y]$ such that $[x]\ne [y]$, we must have $x\not\sim y$, and by part (iii) of the previous
proposition, we deduce that $[x]\cap [y]=\emptyset$. This shows that $A/\sim$ satisfies the second
part of the definition of partition. Lastly, we note that $\emptyset\notin A/\sim$, since
every element of $A/\sim$ is equal to $[x]$ for some $x\in A$, and must thus contain at least
the element $x$.\slug

Let us now revisit the examples of equivalence relations from last class, and see what partitions
they give rise to. In the example $F$ of propositional formulas with variables $p$, $q$, and $r$,
under the equivalence relation $\equiv$, the set of equivalence classes is the set of all possible
truth tables on three variables. Each such truth table has $8$ rows, so there are $2^8$ equivalence
classes. In other words, $|F/\equiv| = 2^8$.

How about the relation $R$ defined on $\RR^2$ where
$$(x,y)\sim (w,z) \qquad\hbox{if and only if}\qquad \sqrt{x^2+y^2} = \sqrt{w^2 + z^2}?$$
Well, each $\bigl[(x,y)\bigr]$ is the circle of radius $\sqrt{x^2 + y^2}$ centred
around the origin, so $\RR^2/\sim$ is the set of all circles in the plane centred at $(0,0)$.

\vfill\eject
\begingroup\headline{\hfil}\footline{\hfil}
\vglue160pt
\centerline{\titlefont II. NUMBER THEORY}
\vskip220pt
\bigskip
\begingroup\obeylines\eightssi
\hfill Die Mathematik ist die K\"onigin der Wissenschaften
\hfill und die Zahlentheorie ist die K\"onigin der Mathematik.
\eightss
\smallskip
\hfill --- {\eightssi Attributed to} C. F. GAUSS {\eightssi (1777--1855)}
\endgroup%\obeylines
\bigskip\goodbreak
\vfill\eject

\advsect Division

Let $a,b\in \ZZ$. We say that $a$ {\it divides} $b$ (or $b$ is a {\it multiple} of $a$, or $b$ is {\it divisible}
by $a$, or $a$ is a {\it factor} of $b$) if there exists $n\in \ZZ$ such that $b=na$. In this
case we write $a\divides b$. For example, $2\divides 10$, since $10 = 5\cdot 2$, but
$3$ does not divide $10$, since there does not exist $n\in \ZZ$ such that $10 = 3n$.
This defines a relation on $\ZZ$.

Note that for all $n\in \ZZ$, $n\divides 0$, since $0 = 0\cdot n$. We also have
$1\divides n$ for all $n\in \ZZ$< since $n = n\cdot 1$. It is true that $0\divides 0$, since we
have, say, $0 = 1\cdot 0$, but for all nonzero $n\in \NN$, $0$ does not divide $n$, since all multiples
of $0$ equal $0$ (and thus cannot equal $n$). Further properties of the ``divides'' relation
are given by the next proposition.

\proclaim Proposition \advthm. For all $a,b,c,d\in \ZZ$,
\medskip
\item{i)} if $a\divides b$, then $a\divides bc$;
\smallskip
\item{ii)} if $a\divides b$ and $a\divides c$, then $a\divides (b+c)$;
\smallskip
\item{iii)} if $a\divides b$ and $b\divides c$, then $a\divides c$;
\smallskip
\item{iv)} if $a\divides b$ and $b\ne 0$, then $|a|\le |b|$; and
\smallskip
\item{v)} if $a\divides b$ and $b\divides a$, then $|a| = |b|$.
\medskip

\proof We leave parts (i) and (ii) as exercises to the reader.

For part (iii), if $a\divides b$ and $b\divides c$,
then there are integers $k$ and $l$ such that $b = ka$ and $c = lb$. Then $c = kla$, so $a\divides c$
(since $kl$ is also an integer).

For part (iv), suppose that $a\divides b$ and $b\ne 0$. Then $b = ka$ for some $k\in \ZZ$,
and $k\ne 0$ since $b\ne 0$. This means that $|b| = |k|\cdot |a|$, but $|k|\ge 1$,
so $|b|\ge |a|$.

For part (v), assume that $a\divides b$ and $b\divides a$. If $a\ne 0$ and $b\ne 0$, then we may apply
part (iv) twice to get $|a|\le |b|$ and $|b|\le |a|$, which together imply $|a| = |b|$. If $b=0$,
then $0\divides a$ so $a = 0$ as well, and $|a| = |b|$. Likewise, if $a = 0$, then $0\divides b$,
so $b=0$ and in this case as well, $|a| = |b|$.\slug

In grade school, you learned how to divide an integer by another one, obtaining a quotient and a remainder.
We state this as a theorem. Its proof (which we shall consider outside the scope of the course,
but which we include as optional reading for those interested)
relies on the well-ordering principle (which we've used
already in these notes)
and the {\it Archimedean property} of $\RR$, which states that for every $x\in \RR$ there
exists $n\in \NN$ such that $n > x$. (One can learn the proof of the Archimedean property from an introductory
course in analysis, e.g., {\mc MATH} 242/254.)

\parenproclaim Theorem {\advthm} (Division algorithm). Let $a,b\in \ZZ$ with $b> 0$. Then there
exist unique integers $q$ and $r$ such that $a = bq+r$ and $0\le r\le |b|$.

\noindent\llap{*}{\it Proof.}\enspace First we show that such integers $q$ and $r$
exist. If $b\divides a$ then $a = kb$ for some $k\in \ZZ$,
and we can set $q = k$ and $r = 0$.

If $a$ is not divisible by $b$, then consider the numbers
$$ \ldots, a - 3b, a-2b, a-b, a, a+b, a+2b, a+3b, \ldots.$$
Let $S$ be the set of these integers that are positive. Symbolically, we have
$$ S = \{a-kb : k\in \ZZ\ \hbox{and}\ a-kb\ge 0\}.$$
By the Archimedean property, there is some $n\in \NN$ such that $n> -a$, which
implies that $nb \ge n > -a$ (here we use the fact that if $b>0$ and $b$ is an integer, then
$b\ge 1$). From this we derive $a > -nb$, and hence $a+nb = a - (-n)b > 0$. This shows
that $S$ is nonempty.

Since $S$ is a nonempty subset of $\NN$, by the well-ordering principle it has a least
element, call it $r$. By definition of $S$, there must be some integer $q$ such that
$r = a-qb$, so $a = bq + r$. We now claim that $0<r<b$.

We know that $r> 0$, since all elements of $S$ are positive by definition. Suppose,
for a contradiction, that $r\ge b$. Then $a - bq = r \ge b$, and so
$$0 \le r-b = a - qb - b = a-(q+1)b.$$
Since $q+1$ is an integer, by definition of $S$, either $r-b$ is an element of $S$,
or $r-b = 0$. Since $r$ was defined to be the minimal element of $S$, it cannot be the
case that $r-b$ is in $S$. So $r-b = 0$. But this means that $0 = a-(q+1)b$; that is,
$a = (q+1)b$, contradicting our assumption that $b$ does not divide $a$. The contradiction
allows us to conclude that $0<r<b$ (in the case that $b$ does not divide $a$). In general,
we have shown that $0\le r<b$.
